Variable Importance for Oblique Learning Ensembles of Trees (VIOLET)

### 2022-03-10 index.Rmd
catg variables added; it seemed necessary

set.seed(329)

ss_demo <-
  sim_surv(
    n_obs = 5000,
    n_z = 3,
    n_x = 3,
    n_g = 3,
    n_w = 3,
    n_v = 3
  )

@@

__Column means__: This test verifies that the mean of each predictor is close to zero, and that roughly 40% of outcomes were censored.

@@

data_means <- apply(data_sim, 2, mean) |> 
  round(3) |> 
  enframe()

@@

data_gt <-  
  benchmark_vi_smry |>
  ungroup() |> 
  mutate(xcorr = recode(xcorr, 
                        '0.0' = '|Corr(linear effects, irrelevant effect)| = 0',
                        '0.1' = '|Corr(linear effects, irrelevant effect)| \u2264 0.10',
                        '0.2' = '|Corr(linear effects, irrelevant effect)| \u2264 0.20',
                        '0.3' = '|Corr(linear effects, irrelevant effect)| \u2264 0.30',
                        '0.4' = '|Corr(linear effects, irrelevant effect)| \u2264 0.40'),
         model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall, x, w, g, v) |> 
  pivot_wider(names_from = n_obs, 
              values_from = c(overall, x, w, g, v)) |> 
  select(xcorr,
         model,
         overall_2500,
         x_2500,
         w_2500,
         g_2500,
         v_2500,
         overall_5000,
         x_5000,
         w_5000,
         g_5000,
         v_5000) |> 
  arrange(xcorr, model)

cols_to_highlight <- c("overall_2500",
                       "x_2500",
                       "w_2500",
                       "g_2500",
                       "v_2500",
                       "overall_5000",
                       "x_5000",
                       "w_5000",
                       "g_5000",
                       "v_5000")

@@

results_tbl_1 <- data_gt |> 
  mutate(across(ends_with("0"), ~table_value(100*.x))) |> 
  gt(rowname_col = 'model', groupname_col = 'xcorr') |> 
  tab_stubhead(label = 'Variable importance technique') |> 
  tab_spanner(label = 'N = 2,500', 
              columns = c("overall_2500",
                          "x_2500",
                          "w_2500",
                          "g_2500",
                          "v_2500")) |> 
  tab_spanner(label = 'N = 5,000', 
              columns = c("overall_5000",
                          "x_5000",
                          "w_5000",
                          "g_5000",
                          "v_5000")) |> 
  cols_label(
    overall_2500 = "All effects",
    x_2500 = "Linear effects",
    w_2500 = "Non-linear effects",
    g_2500 = "Interaction effects",
    v_2500 = "Combo effects",
    overall_5000 = "All effects",
    x_5000 = "Linear effects",
    w_5000 = "Non-linear effects",
    g_5000 = "Interaction effects",
    v_5000 = "Combo effects"
  ) |> 
  cols_align(align = 'center')

@@

@@

data_fig <- benchmark_vi_smry |>
  ungroup() |> 
  mutate(model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall) |> 
  group_by(n_obs, model, xcorr) |> 
  summarize(value = mean(overall))

ggplot(data_fig) + 
  aes(x = xcorr, y = value, col = model) + 
  geom_line() +
  facet_wrap(~n_obs)

results_tbl_2 <- benchmark_vi_smry |>
  ungroup() |> 
  mutate(model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall, x, w, g, v) |> 
  group_by(model) |> 
  summarize(
    across(
      c(overall, x, w, g, v),
      ~table_value(mean(.x))
    )
  ) |> 
  gt(rowname_col = 'model') |> 
  tab_stubhead(label = 'Variable importance technique') |> 
  cols_label(
    overall = "Overall",
    x = "Linear effects",
    w = "Non-linear effects",
    g = "Interaction effects",
    v = "Combo effects"
  ) |> 
  cols_align(align = 'center')

>>>>

set.seed(32987)

ss_demo <-
  sim_surv(
    n_obs = 5000,
    n_z = 3,
    n_x = 3,
    n_g = 3,
    n_w = 3,
    n_v = 3,
    # omit categorical variables; these are tested separately
    n_c = 0 
  )

@@

__Column means__: This test verifies that the mean of each predictor is close to zero, and that roughly 45% of outcomes were censored.

@@

data_means <- data_sim |> 
  select(where(is.numeric)) |> 
  apply(2, mean) |> 
  round(3) |> 
  enframe()

@@

data_gt <-  
  benchmark_vi_smry |>
  ungroup() |> 
  mutate(xcorr = recode(xcorr, 
                        '0.0' = '|Corr(linear effect variables, irrelevant variables)| = 0',
                        '0.1' = '|Corr(linear effect variables, irrelevant variables)| \u2264 0.10',
                        '0.2' = '|Corr(linear effect variables, irrelevant variables)| \u2264 0.20',
                        '0.3' = '|Corr(linear effect variables, irrelevant variables)| \u2264 0.30',
                        '0.4' = '|Corr(linear effect variables, irrelevant variables)| \u2264 0.40'),
         model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall, x, w, g, v, c) |> 
  pivot_wider(names_from = n_obs, 
              values_from = c(overall, x, w, g, v, c)) |> 
  select(xcorr,
         model,
         overall_2500,
         x_2500,
         w_2500,
         g_2500,
         v_2500,
         c_2500,
         overall_5000,
         x_5000,
         w_5000,
         g_5000,
         v_5000,
         c_5000) |> 
  arrange(xcorr, model)

cols_to_highlight <- c("overall_2500",
                       "x_2500",
                       "w_2500",
                       "g_2500",
                       "v_2500",
                       "c_2500",
                       "overall_5000",
                       "x_5000",
                       "w_5000",
                       "g_5000",
                       "v_5000",
                       "c_5000")

@@

results_tbl_1 <- data_gt |> 
  mutate(across(ends_with("0"), ~table_value(100*.x))) |> 
  gt(rowname_col = 'model', groupname_col = 'xcorr') |> 
  tab_stubhead(label = 'Variable importance technique') |> 
  tab_spanner(label = 'N = 2,500', 
              columns = c("overall_2500",
                          "x_2500",
                          "w_2500",
                          "g_2500",
                          "v_2500",
                          "c_2500")) |> 
  tab_spanner(label = 'N = 5,000', 
              columns = c("overall_5000",
                          "x_5000",
                          "w_5000",
                          "g_5000",
                          "v_5000",
                          "c_5000")) |> 
  cols_label(
    overall_2500 = "All effects",
    x_2500 = "Linear effects",
    w_2500 = "Non-linear effects",
    g_2500 = "Interaction effects",
    v_2500 = "Combo effects",
    c_2500 = "Categorical effects",
    overall_5000 = "All effects",
    x_5000 = "Linear effects",
    w_5000 = "Non-linear effects",
    g_5000 = "Interaction effects",
    v_5000 = "Combo effects",
    c_5000 = "Categorical effects"
  ) |> 
  cols_align(align = 'center')

@@

data_fig <- benchmark_vi_smry |>
  ungroup() |> 
  mutate(
    n_obs = recode(n_obs,
                   '2500' = 'Training data sample size = 2,500',
                   '5000' = 'Training data sample size = 5,000'),
    model = factor(model,
                   levels = c('aorsf',
                              'aorsf_menze',
                              'xgboost',
                              'randomForestSRC'),
                   labels = c('Negation importance',
                              'ANOVA importance',
                              'SHAP importance',
                              'Permutation importance'))
  )

fig <- ggplot(data_fig) + 
  aes(x = xcorr, y = overall, col = model) + 
  geom_line() +
  geom_point(size = 2) +
  scale_color_manual(values = c("purple", "orange", "grey", "limegreen")) +
  facet_wrap(~n_obs) + 
  labs(x = "max( |Corr(linear effect variables, irrelevant variables)| )",
       y = "Discrimination between relevant\nand irrelevant predictor variables",
       color = 'Variable importance\nestimated using') + 
  theme_bw() + 
  theme(panel.grid = element_blank())

ggsave(filename = 'plot_violet.png',
       plot = fig,
       device = 'png',
       width = 9,
       height = 4,
       dpi = 600)

fig

@@

results_tbl_2 <- benchmark_vi_smry |>
  ungroup() |> 
  mutate(model = recode(model,
                        'aorsf' = 'Negation importance',
                        'aorsf_menze' = 'ANOVA importance',
                        'xgboost' = 'SHAP importance',
                        'randomForestSRC' = 'Permutation importance')) |> 
  select(n_obs, model, xcorr, overall, x, w, g, v, c) |> 
  group_by(model) |> 
  summarize(
    across(
      c(overall, x, w, g, v, c),
      ~table_value(mean(.x))
    )
  ) |> 
  gt(rowname_col = 'model') |> 
  tab_stubhead(label = 'Variable importance technique') |> 
  cols_label(
    overall = "Overall",
    x = "Linear effects",
    w = "Non-linear effects",
    g = "Interaction effects",
    v = "Combo effects",
    c = "Categorical effects"
  ) |> 
  cols_align(align = 'center')
